---
title: '732A90: Computational Statistics'
author: "Sofie Jörgensen, Oriol Garrobé Guilera, David Hrabovszki"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: my_header.tex
  html_document:
    df_print: paged
subtitle: Computer lab6 - Group11
header-includes: \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r warning = FALSE}
# R version
RNGversion('3.5.1')
```

# Question 1: Genetic algorithm

In this exercise we are going to perform one-dimensional maximization by using a genetic algorithm.

### 1.

Firstly, we define the function \texttt{f()} as

$$
f(x):= \frac{x^2}{e^x} -2 \ \text{exp} (-(9\text{sin } x)/(x^2+x+1)). 
$$

```{r}
#1.1
f <- function(x){
  return(x^2/exp(x) - 2*exp(-1*(9*sin(x)) / (x^2 + x + 1)))
}

```


### 2.

Secondly, we define the function \texttt{crossover()}, that takes two scalars $x$ and $y$ as inputs, and returns a child as $\frac{x+y}{2}$.

```{r}
#1.2
crossover <- function(x,y){
  return((x+y) / 2)
}

```

### 3.

Thirdly, we define the function \texttt{mutate()}, that performs the integer division $x^2 \text{  mod } 30$, for a scalar input $x$. 

```{r}
#1.3
mutate <- function(x){
  return(x^2 %% 30)
}
```


### 4.


Further, we will create a function called \texttt{genetic()}, with the parameters \texttt{maxiter} and \texttt{mutprob}. The settings of this function as well as the results are presented in (a)-(e). The code can be found in the Appendix.


(a). The function $f()$ is plotted in the range from 0 to 30 in Figure X, and we can observe that there is a maximum value at ...

(b)


### 5.






# Question 2: EM algorithm


### 1.





### 2.





### 3.





### 4.








```{r warning=FALSE}
# R version
RNGversion('3.5.1')
# Packages

```



----------

# Appendix

```{r ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}

```